Demystifying the Computer Science PhD Admission in the US\\{\large A Guide for International Students

### 2023-11-25 demystify.tex
update

\begin{domesticbox}[Domestic students:] If you're a domestic student, you actually have several advantages in your application.  First, we already know about your school (see \S\ref{sec:your-school}), and you are also more familiar with the US education system and culture.  You also have more opportunities for funding, e.g., through government scholarships for US citizens and residents. Finally, compared to international students, PhD applications in CS from domestic students are also few and that can help your case.
\end{domesticbox}

@@

\section{Graduating Assistantship (TA/RA)}
The most common type of funding is \textbf{graduate assistanship}, which is either TA or RA. Both TA and RA come with tuition waiving (you don't have to pay tuition), health insurance (this takes care of your insurance, which is a must have in the US), and most importantly, your stipend (i.e., your salary). Some universities also pay insurance for spouse/children (or give very good discount).

@@

Students often complain their stipend being too low, but it is actually not bad and you can live comfortably with it.  In many cases, it is enough to support your spouse and kids (many CS PhD students have their spouse and kids with them). So don't worry too much about stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive, otherwise it will loose good students to other schools.  For example, at GMU, every year we discuss about improving the benefits, and especially stipend, for our graduate students.

@@

Choosing a school and an adviser is clearly one of the most important things in your mind when you apply and especially when you get admitted.  This is further complicated due to cultural differences and unfamiliarity of  international students to the US higher education system.  This section aims to mitigate some confusions and help you make informed decision.

@@

Ultimately, choose one that fits you the most by communicating with them, taking their courses, meeting them, asking them questions, and even talking to their current students. Indeed all this takes a lot of time and effort, but remember, you will be working with this person for 4-5 years, so it is important to try to find the right one.

@@

\subsection{Tenured, Tenure Track, Teaching,  Assistant Profs., Associate Profs}\label{sec:faculty-types}

@@

% \begin{commentbox}[Thanh:]
% \end{commentbox}

@@

\section{Differences between PhD in the US and other countries}\label{sec:non-us-differences}
This summarizes the main differences between CS PhD in the US and other countries. % This is based on my experience and what I heard from others who did PhD in other countries.

@@

\emph{Work-life balance}: I believe that PhD students are often overworked in the US compared to other countries, e.g., in Europe.  This is partly due to the longer PhD program and that US PhD students are often paid through TA, which requires them to do TA in addition to their own research. In contrast, PhD students in other countries are often paid through fellowships, which do not require them to do TA work.

@@

If you're reaching out to a professor for the first time,  address them as Prof. Lastname or Dr. Lastname. Many international students use Prof. or Dr. FirstName LastName, but this can come across as if you're simply copying and pasting names. It's not necessary, so stick with Prof. or Dr. Lastname.

>>>>

\begin{domesticbox}[Domestic students:] If you're a domestic student, you have several advantages in your application.  First, we already know about your school (see \S\ref{sec:your-school}), and you are also more familiar with the US education system and culture.  You also have more opportunities for funding, e.g., through government scholarships for US citizens and residents. Finally, compared to international students, PhD applications in CS from domestic students are few and that can help your case.
\end{domesticbox}

@@

\section{Graduate Assistantship (TA/RA)}
The most common type of funding is \textbf{graduate assistantship}, which is either TA or RA. Both TA and RA come with tuition waiving (you don't have to pay tuition), health insurance (this takes care of your insurance, which is a must have in the US), and most importantly, your stipend (i.e., your salary). Some universities also pay insurance for spouse/children (or give very good discount).

@@

Students often complain their stipend being too low, but it is actually not bad and you can live comfortably with it.  In many cases, it is also enough to support your spouse and kids (many CS PhD students have their family with them). So don't worry too much about stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at GMU, every year we discuss about improving the benefits, and especially stipend, for our graduate students.

@@

Choosing a school and an adviser is clearly among the most important things in your mind when you apply and especially when you get admitted.  This is further complicated due to cultural differences and unfamiliarity of  international students to the US higher education system.  This section aims to mitigate some confusions and help you make informed decision.

@@

Ultimately, choose a prof. that fits you the most by communicating with them, taking their courses, meeting them, asking them questions, and even talking to their current students. Indeed all this takes a lot of time and effort, but remember, you will be working with this person for 4-5 years, so it is important to try to find the right one.

@@

\subsection{Tenured, Tenure Track, Teaching,  Assistant Profs., Associate Profs, and O
ther Fancy Academic Titles}\label{sec:faculty-types}

@@

@@

\section{Differences between PhD in the US and Other Countries}\label{sec:non-us-differences}

This summarizes the main differences between CS PhD in the US and other countries. % This is based on my experience and what I heard from others who did PhD in other countries.

@@

\emph{Work-life balance}: I believe that PhD students in the US are overworked compared to other countries, e.g., in Europe.  This is partly due to the longer PhD program and that US PhD students are often paid through TA, which requires them to do TA in addition to their own research. In contrast, PhD students in other countries are often paid through fellowships, which do not require them to do TA.

@@

If you're reaching out to a professor for the first time,  address them as Prof. or Dr. Lastname. Many international students use Prof. or Dr. FirstName LastName, but this can come across as if you're simply copying and pasting names. It's not necessary, so stick with Prof. or Dr. Lastname.
